\bf Hierarchical Network Structure Promotes Dynamical Robustness

### 2014-11-26 paper.tex
move supplement to main text
move figures into main text and switch bibliography format
update text

% Title must be 150 characters or less
\begin{center}
{\Large
\textbf{Hierarchical Network Structure Promotes Dynamical Robustness}
}
% Insert Author names, affiliations and corresponding author email.
\\[.5cm]
Cameron Smith$^{1}$,
Raymond S. Puzio$^{1}$,
Aviv Bergman$^{1,2,3,4}$,
\\[.5cm]
$^1$Department of Systems and Computational Biology,\\
$^2$Dominick P. Purpura Department of Neuroscience,\\
$^3$Department of Pathology, Albert Einstein College of Medicine,\\
1301 Morris Park Ave, Bronx, NY 10461, USA\\
$^4$Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, NM 87501, USA
\\[.5cm]
% $\ast$To whom correspondence should be addressed; E-mail: aviv@einstein.yu.edu.
\end{center}

{\begin{quote} \bf
\input{tex/abstract-cited}
\end{quote}}

@@

\section{Acknowlegements}  The authors would like to thank Jay Sulzberger and Ximo Pechuan for useful suggestions and discussions.

\pagebreak

\bibliographystyle{pnas}
\bibliography{bib/books,bib/papers}

@@

\section{Figure Legends}
\input{tex/figures}

\beginsupplement
\setcounter{secnumdepth}{4}
\section{Supplementary Material}

\input{tex/supplement}
\pagebreak

@@

% \section{Tables}
% \todo{Fill out notation table}
% \input{tex/tablenotation}
\input{tex/table2x2stab}
\pagebreak
\input{tex/tablenographs}

% \input{tex/tablewithgraphs}
% \input{tex/extra}

@@

%------------------------------------------------------------------------------
% End of journal.tex
%------------------------------------------------------------------------------

>>>>

\title{\bf Hierarchical Network Structure Promotes Dynamical Robustness}

\author{Cameron Smith$^{1}$,
Raymond S. Puzio$^{1}$,
Aviv Bergman$^{1,2,3,4,*}$}

\affiliation{$^1$Department of Systems and Computational Biology,\\
$^2$Dominick P. Purpura Department of Neuroscience,\\
$^3$Department of Pathology, Albert Einstein College of Medicine,\\
1301 Morris Park Ave, Bronx, NY 10461, USA\\
$^4$Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, NM 87501, USA
\\
$*$to whom correspondence should be addressed: aviv@einstein.yu.edu}

\date{\today}

\input{tex/abstract-cited}

\maketitle

@@

\section{Methods}
{\footnotesize
\input{tex/methods}
}

\section{Acknowlegements}
\input{tex/acknowledgements}

\bibliographystyle{unsrtnat}
\bibliography{bib/books,bib/papers}

\onecolumngrid

@@

\beginsupplement
\setcounter{secnumdepth}{4}
\section{Supporting Information}

@@

% \section{Tables}
\input{tex/table2x2stab}
\pagebreak
\input{tex/tablenographs}

@@

%------------------------------------------------------------------------------
% End of paper.tex
%------------------------------------------------------------------------------

### 2014-11-26 tex/abstract-cited.tex
update text

%!TEX root = ../paper-abstract.tex
The analysis of dynamical systems that attempts to model chemical reaction, gene-regulatory, population, and ecosystem networks all rely on models having interacting components~\cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}. When the details of these interactions are unknown for biological systems of interest, one effective approach is to study the dynamical properties of an ensemble of models determined by evolutionary constraints that may apply to all such systems~\cite{Gardner1970,May1972,Cohen1984,May1972a,Radius2014}. One such constraint is that of dynamical robustness, the probability of a stable system to remain stable under perturbation~\cite{WADDINGTON1942a,Wagner1997,Rutherford1998,VanNimwegen1999,Siegal2002,Bergman2003,Ciliberti2007b,Ciliberti2007,Draghi2010,Wagner2013}. Despite previous investigations, the relationship between dynamical robustness---an important functional characteristic of many biological systems---and network structure is poorly understood. Here we analyze the stability and robustness of a large class of dynamical systems and demonstrate that the most hierarchical network structures are the most robust. In particular, we determine the probability distribution of robustness over system connectivity and show that robustness is maximized by maximizing the number of links between strongly connected components of the graph representing the underlying system connectivity. We also demonstrate that permutation of strongly connected components is a fundamental symmetry of dynamical robustness, which applies to networks of any number of components.  The classification of dynamical robustness based upon a purely topological property provides a fundamental organizing principle that can be used in the context of experimental validation to select among models that break or preserve network hierarchy. This result contributes to an explanation for the observation of hierarchical modularity in biological networks at all scales~\cite{Zhao2006,Ravasz2002,Bhardwaj2010,Colm,Corominas-Murtra2013}.

>>>>

%!TEX root = ../paper.tex
\begin{abstract}
The analysis of dynamical systems that attempts to model chemical reaction, gene-regulatory, population, and ecosystem networks all rely on models having interacting components~\cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}. When the details of these interactions are unknown for biological systems of interest, one effective approach is to study the dynamical properties of an ensemble of models determined by evolutionary constraints that may apply to all such systems~\cite{Gardner1970,May1972,Cohen1984,May1972a,Radius2014}. One such constraint is that of dynamical robustness, the probability of a stable system to remain stable under perturbations to the strength of interactions among its components~\cite{WADDINGTON1942a,Wagner1997,Rutherford1998,VanNimwegen1999,Siegal2002,Bergman2003,Ciliberti2007b,Ciliberti2007,Draghi2010,Wagner2013}. Despite previous investigations, the relationship between dynamical robustness---an important functional characteristic of many biological systems---and network structure is poorly understood. Here we analyze the stability and robustness of a large class of dynamical systems and demonstrate that the most hierarchical network structures are the most robust. In particular, we determine the probability distribution of robustness over system connectivity and show that robustness is maximized by maximizing the number of links between strongly connected components of the graph representing the underlying system connectivity. We demonstrate that this can be understood in terms of the fact that permutation of strongly connected components is a fundamental symmetry of dynamical robustness, which applies to networks of any number of components and is independent of the distribution from which the strengths of interconnection among components are sampled.  The classification of dynamical robustness based upon a purely topological property provides a fundamental organizing principle that can be used in the context of experimental validation to select among models that break or preserve network hierarchy. This result contributes to an explanation for the observation of hierarchical modularity in biological networks at all scales~\cite{Zhao2006,Ravasz2002,Bhardwaj2010,Colm,Corominas-Murtra2013}.
\end{abstract}

### 2014-11-26 tex/introduction.tex
update text

Recent work has demonstrated that as a result of inhomogeneity in the geometry of the sensitivity of the dynamics over the parameter space for such models, this approach allows for a large variety of models to fit the data \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014,Tonsing2014}. In addition, there is often uncertainty about the very structure of such biological networks. In this context, it is crucial to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems, which is necessary to understand in order to determine whether or not a particular dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}. This can be achieved using a method common in statistical physics involving the consideration of an ensemble of systems that, in comparison to one another, appear to have components that are randomly interlinked.

Indeed, investigating generic properties of a large class of dynamical systems was the approach taken by May in models of ecosystem dynamics \cite{Gardner1970,May1972}. The class of dynamical systems studied by May is, however, not restricted to ecosystem dynamics and encompasses, among others, the dynamics of all of the biological networks represented in \ref{fig:biomodelexamples}. May conjectured on the basis of results from random matrix theory what eventually came to be referred to as the May-Wigner stability theorem \cite{Cohen1984,May1972a,Radius2014}, which implies a relationship between a topological property, system connectivity, and a dynamical property, stability.

% The generic applicability of gaining a better understanding of the class of models investigated by May demonstrates the unequivocal value of deeper investigation.  However, work attempting to continue the development of the so-called May-Wigner stability theorem revealed that May's conjectured stability criterion was not as easy to demonstrate as was initially believed.

@@

>>>>

Recent work has demonstrated that as a result of inhomogeneity in the sensitivity of the dynamics over the parameter space for such models, this approach allows for a large variety of models to fit the data \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014,Tonsing2014}. In addition, there is often uncertainty about the very structure of such biological networks. In this context, it is crucial to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems, which is necessary to understand in order to determine whether or not a particular dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}. This can be achieved using a method common in statistical physics involving the consideration of an ensemble of systems that, in comparison to one another, appear to have components that are randomly interlinked.

Indeed, investigating generic properties of a large class of dynamical systems was the approach taken by May in models of ecosystem dynamics \cite{Gardner1970,May1972}. The class of dynamical systems studied by May is, however, not restricted to ecosystem dynamics and encompasses, among others, the dynamics of all of the biological networks represented in \ref{fig:biomodelexamples}. May conjectured on the basis of results from random matrix theory what eventually came to be referred to as the May-Wigner stability theorem \cite{Cohen1984,May1972a,Radius2014,Majumdar2014}, which implies a relationship between a topological property, system connectivity, and a dynamical property, stability.

Some conclusions derived from this approach may be sensitive to the probability distribution from which the strengths of interconnection are sampled, whereas others are independent of the specific form of this distribution. In the former case, discovering the constraints that contribute to determining the functional form of the distribution can be a matter for empirical investigation and results that are themselves functions of this distribution can be reinterpreted in light of ongoing updates from relevant empirical observations. On the other hand, results that are independent of the specific form of this distribution can be expected to hold in any case and a specific distribution may simply be chosen for the purposes of exemplifying the result.

@@

Here we determine the relationship between network hierarchy, a topological property, and the probability of \emph{robustness}, a dynamical property. Robustness is of interest to biological systems at all scales, and has been previously studied in the context of gene-regulatory networks \cite{WADDINGTON1942a,VanNimwegen1999,Siegal2002,Ciliberti2007b,Ciliberti2007,Wagner2013}. Intuitively, robustness is the probability of stability to perturbation in the system structure or its parameters for systems which have already been determined to be stable in the sense of linear stability analysis \cite{Davis1962}. We demonstrate that systems with the most hierarchical network topology exhibit maximal robustness and explain why this results from an invariance of robustness to particular kinds of transformations of the network topology.
